---
title: "Milestone #3"
author: "Group 22: Jaemie Anne Abad, Vig Karthik, Kathy LeBert"
date: "2023-11-06"
output: html_document
---

## Code Chunks from Milestone #2

Milestone #2 Summary
Our project continues with the two Cal Enviro Screen (CES) datasets from 2021 ("scores" and "measures") and the California Health and Human Services (CHHS) asthma emergency department (ED) visit rates dataset from 2020.

As a summary of what was accomplished in the previous milestone, we imported the three datasets into R, renamed the column names for ease of analysis, and identified some variables of interest to use in our analysis. We ran summary statistics in our initial exploration of the data.
The datasets contain rows of observations by census tract, which are further categorized into counties. While this level of granularity in location is appreciated, our team wishes to match asthma ED visit rates with the CES scores. The asthma ED visit rates are only at the county level, leading to the need to aggregate census tracts into the county level to match up the datasets for further analysis.


```{r clean up columns - Vig}

# update.packages()

library(data.table)
library(readxl)
library(tidyverse)

#loading the data.table library to read in the csv
#also loaded read_xl to read the xlsx file

data_dictionary <- read_xlsx("calenviroscreen_datadictionary.xlsx")
#using read_xlsx to read in the data dictionary

measures_raw <- fread("calenviroscreen_measures_2021.csv", encoding = "UTF-8")
#using fread() from the data.table package to read the measures csv.
#using fread() so special characters read in properly

scores_raw <- fread("calenviroscreen_scores_demog_2021.csv", encoding = "UTF-8")
#using fread() to read the demographics csv

asthma_raw <- fread("chhs_asthma_ed.csv")
#using fread() to read the asthma csv.

#now that we've read the files, I've went into the csv files and removed 
#periods by some abbreviations to avoid '._' string chunks in columns
#this is mostly for aesthetic and typing ease during the project

#the below functions take the raw csvs and convert all columns to lowercase
#they also remove spaces in the column names to convert columns to snakecase

measures <- rename_with(
  measures_raw,             # data frame
  ~ tolower(           # function call (using tolower())
    gsub(" ",          # embedded gsub() checking for empty spaces pattern " "
         "_",          # replace pattern with underscore "_"         
         .x,           
         fixed = TRUE) 
    ))

scores <- rename_with(
  scores_raw,             # data frame
  ~ tolower(           # function call (using tolower())
    gsub(" ",          # embedded gsub() checking for empty spaces pattern " "
         "_",          # replace pattern with underscore "_"         
         .x,           
         fixed = TRUE) 
    ))


asthma <- rename_with(
  asthma_raw,             # data frame
  ~ tolower(           # function call (using tolower())
    gsub(" ",          # embedded gsub() checking for empty spaces pattern " "
         "_",         # replace pattern with underscore "_"          
         .x,           
         fixed = TRUE) 
    ))

str(measures)
str(scores)
str(asthma)

#I checked the dataframe structure to see if we needed to change column types.
#The column types seem to be accurately assigned. Based on the column
#descriptors, there seems to be no mismatch of data types. Characters are
#characters, integers are integers, and decimals are numbers.

```

```{r data_element_analysis - Jaemie}

# # 1 - CES County
# # lists which of CA's 58 counties is specified
# class(measures$california_county)
# 
# # 2 - CES Asthma
# class(measures$asthma)
# summary(measures$asthma)
# class(measures$asthma_pctl)
# summary(measures$asthma_pctl)
# 
# # 3 - CES Score
# class(scores$ces_4.0_score)
# summary(scores$ces_4.0_score)
# class(scores$ces_4.0_percentile)
# summary(scores$ces_4.0_percentile)
# 
# # 4 - CES Age
# class(scores$children_below_10_years_prct)
# summary(scores$children_below_10_years_prct)
# class(scores$pop_10_to_64_years_prct)
# summary(scores$pop_10_to_64_years_prct)
# class(scores$elderly_above_64_years_prct)
# summary(scores$elderly_above_64_years_prct)
# 
# # 5 - CES Race
# class(scores$white_prct)
# summary(scores$white_prct)
# class(scores$hispanic_prct)
# summary(scores$hispanic_prct)
# class(scores$african_american_prct)
# summary(scores$african_american_prct)
# class(scores$native_american_prct)
# summary(scores$native_american_prct)
# class(scores$asian_american_prct)
# summary(scores$asian_american_prct)
# class(scores$`other/multiple_prct`)
# summary(scores$`other/multiple_prct`)
# 
# # 6 - CHHS Asthma
# class(asthma$year)
# summary(asthma$year)
# class(asthma$`age-adjusted_ed_visit_rate`)
# summary(asthma$`age-adjusted_ed_visit_rate`)
# # also include county

```

## Data Subsetting, Cleaning, and Pivoting

Data Subsetting, Cleaning, and Pivoting

To start with, we first subsetted the asthma ED visit rates dataset, called "asthma", to the year 2020. The dataset contains other years, but we have chosen to focus on the year 2020 as it is the nearest year to the CES datasets, which contain data from 2021.

Next, we converted the census tracts in the CES datasets from the integer class to the character class, as they are number-based location IDs. We then identified the missing values in the three datasets and calculated the proportion of the missing values for each. For example, the "asthma" dataset contains 4.5% missing values within any of its columns, while "measures" contains 0.7%, and "scores" contains 0.4%. These numbers seem to be low, but our team will conduct further research and discussions to figure out how to handle these missing values.

The "scores" dataset only contained race and age data in the form of percentages, but luckily also had the total population for each census tract. We were able to convert these race and age percentages into population numbers, and then rounded to the nearest whole number. We also rounded CES scores and their corresponding percentiles to the nearest 2 decimal places.

We decided that pivoting the data is not needed at this time, as each row already corresponds to a unique census tract.


```{r subset, clean, pivot}

# subset asthma data set by year 2020
asthma_2020 <- subset(asthma, year == 2020)

```
```{r Jaemie 2}
# convert census tract in two datasets from integer to character
class(measures$census_tract)
measures$census_tract <- as.character(measures$census_tract)
class(measures$census_tract)
class(scores$census_tract)
scores$census_tract <- as.character(scores$census_tract)
class(scores$census_tract)

# identify missing values in the three datasets
asthma_missing <- colSums(is.na(asthma))
print(asthma_missing)
measures_missing <- colSums(is.na(measures))
print(measures_missing)
scores_missing <- colSums(is.na(scores))
print(scores_missing)

# what is the proportion of missing values in each dataset?
# https://thomasadventure.blog/posts/r-count-na/
mean(is.na(asthma))
mean(is.na(measures))
mean(is.na(scores))

# rename any more columns as needed
scores <- scores %>%
  rename(other_multiple_prct = `other/multiple_prct`)

```

```{r Population Numbers}
# change race and age percentages to population numbers, round to whole numbers
scores$children_below_10_years_num <- round(scores$total_population * scores$children_below_10_years_prct)
scores$pop_10_to_64_years_num <- round(scores$total_population * scores$pop_10_to_64_years_prct)
scores$elderly_above_64_years_num <- round(scores$total_population * scores$elderly_above_64_years_prct)
scores$hispanic_num <- round(scores$total_population * scores$hispanic_prct)
scores$white_num <- round(scores$total_population * scores$white_prct)
scores$african_american_num <- round(scores$total_population * scores$african_american_prct)
scores$native_american_num <- round(scores$total_population * scores$native_american_prct)
scores$asian_american_num <- round(scores$total_population * scores$asian_american_prct)
scores$other_multiple_num <- round(scores$total_population * scores$other_multiple_prct)

# round to 2 decimal points for CES scores and percentiles
scores$ces_4.0_score_round <- round(scores$ces_4.0_score, digits = 2)
scores$ces_4.0_percentile_round <- round(scores$ces_4.0_percentile, digits = 2)

# pivoting is not needed for these datasets at this time


```

Creating Variables for Analysis
In order to convert census tract data into county-level observations, our team decided to equate counties into means of the census tracts within.

As we wished to analyze any racial health disparities within asthma visit rates and environmental exposures, we looked at the available racial data and decided to form two new variables: non_white_pop, referring to the number of people in each county that are White, and non_white_pop, which is the number of people in each county that are People of Color (non-White). We also wanted to identify any age-related differences, leading to the creation of pop_under_10, which are children below the age of 10, and pop_over_10, which are individuals above the age of 10. These age-related groups were chosen as they are based on the groupings in the CES datasets. 

We hope that the analysis of racial disparities or age-related disparities can provide good evidence for any environmental health interventions to be formed in the near future.


```{r New Variables and Cleaning Part 1}
#combining CES scores into a county level measure

scores_raw_county <- scores_raw %>%
  replace(is.na(.),0) %>%
  select(-c(`CES 4.0 Percentile Range`,`CES 4.0 Percentile`, `Census Tract`)) %>%
  group_by(County) %>% 
  summarize_all(mean) %>%
#lumping together non-white percentages in population
  mutate(non_white_prct =`Hispanic prct`+`African American prct`+`Native American prct`+`Asian American prct`+`Other/Multiple prct`) %>%
#calculating non-white population
  mutate(non_white_pop = `non_white_prct`/100 * `Total Population`) %>%
#calculating white population
  mutate(white_pop = `White prct`/100 * `Total Population`) %>%
  rename(white_prct =`White prct`) %>%
#removing individual race categories for non-white individuals
  select(-c(`Hispanic prct`,`African American prct`,`Native American prct`,`Asian American prct`,`Other/Multiple prct`)) %>%
#calculating population of children below 10 years
  mutate(pop_under_10 = `Children below 10 years prct`/100*`Total Population`) %>%
  rename(pop_under_10_prct = `Children below 10 years prct`) %>%
#calculating population of individuals above 10 years
  mutate(pop_over_10_prct = `Pop 10 to 64 years prct` + `Elderly above 64 years prct`) %>%
  mutate(pop_over_10 = `pop_over_10_prct`/100 * `Total Population`) %>%
#removing sub-categories of individuals over 10 years
  select(-c(`Pop 10 to 64 years prct`,`Elderly above 64 years prct`))

```

We re-calculated the quartiles for the CES datasets, now based on county data instead of census tract data. We selected traditional quartiles in 25% increments. These quartiles will allow us to rank the county data and see where the environmental health burden is the highest and lowest, again providing good evidence for later interventions to be created.


```{r Cleaning Part 2}
 
#calculating quartiles for county CES scores
CES_score_quartile <- quantile(scores_raw$`CES 4.0 Score`, na.rm=TRUE)

scores_raw_county <- scores_raw_county %>%
  mutate(CES_score_quartile_range = cut(`CES 4.0 Score`, breaks = c(-Inf, 14.786523, 25.553718, 40.057395, Inf),
                                        labels = c("Lowest Quartile","Second Quartile","Third Quartile", "Fourth Quartile"))) %>%

#subsetting and ordering the scores dataset
   select(c(`County`, `CES 4.0 Score`, `CES_score_quartile_range`, `Total Population`, `white_prct`, `white_pop`, `non_white_prct`, `non_white_pop`, `pop_under_10_prct`, 
           `pop_under_10`,`pop_over_10_prct`, `pop_over_10`))

scores_raw_county <- scores_raw_county[order(scores_raw_county$County),]
scores_raw_county$County <- toupper(scores_raw_county$County)

```

We recoded invalid values in the datasets and then summarized the means of county measures, again to transform the data into meaningful county-level observations. This also aided in the joining of multiple datasets based on the lowest common level of detail available (county). 

We filtered, stratified, and created datasets based on our exposures of interest: age (child vs. adult), race (White vs. Non-White), and also summarized means of two CES measures of interest (Pollution and Housing Burden). 


```{r Cleaning Part 3}
#recoding invalid values in the measures file
#then summarizing means of county measures (Pollution/Housing Burden)
measures_raw_county <- measures_raw %>%
  replace(is.na(.),0) %>%
  group_by(`California County`) %>%
  summarize_all(mean) %>%
  select(c(`California County`,`Pollution Burden`,`Pollution Burden Score`, `Pollution Burden Pctl`, `Housing Burden`, `Housing Burden Pctl`))
  
measures_raw_county <- measures_raw_county[order(measures_raw_county$`California County`),]
colnames(measures_raw_county)[colnames(measures_raw_county)== "California County"] = "County" 
measures_raw_county$County <- toupper(measures_raw_county$County)
measures_raw_county <- measures_raw_county %>% mutate(County=paste(County,c("COUNTY"), sep=" "))

```


```{r Cleaning Part 4}
#recoding invalid values for asthma dataset
#grouping values by county
asthma_2020 <- asthma %>%
  replace(is.na(.),0) %>%
  group_by(county) %>%
  filter(year>=2020)
  #select(-c(`age-adjusted_ed_visit_rate`, `year`))

#filtering asthma dataset to stratify by age
#create two datasets - one for child, one for adult
asthma_2020_age <- asthma_2020 %>% filter(strata =="Child vs. adult")

asthma_2020_age_adult <- filter(asthma_2020_age, strata_name == "18+ years")
asthma_2020_age_adult <- asthma_2020_age_adult[-(1),]
asthma_2020_age_adult <- asthma_2020_age_adult[order(asthma_2020_age_adult$county),]
colnames(asthma_2020_age_adult)[colnames(asthma_2020_age_adult)== "county"] = "County" 
asthma_2020_age_adult <- asthma_2020_age_adult %>% mutate(County=paste(County,c("COUNTY"), sep=" ")) %>%
  select(c("County","number_of_ed_visits", "age-adjusted_ed_visit_rate")) %>%
  rename(adult_asthma_ed_visits = "number_of_ed_visits")

asthma_2020_age_child <- asthma_2020_age[-(60:118),]
asthma_2020_age_child <- asthma_2020_age_child[-(1),]
asthma_2020_age_child <- asthma_2020_age_child[order(asthma_2020_age_child$county),]
colnames(asthma_2020_age_child)[colnames(asthma_2020_age_child)== "county"] = "County" 
asthma_2020_age_child <- asthma_2020_age_child %>% mutate(County=paste(County,c("COUNTY"), sep=" "))%>%
  select(c("County","number_of_ed_visits", "age-adjusted_ed_visit_rate")) %>%
  rename(child_asthma_ed_visits = "number_of_ed_visits")



#filtering asthma dataset to stratify by race
#create two datasets - one for white, one for non-white
#summarizing number of ED visits for white and non-white individuals by county

asthma_2020_race <- asthma_2020 %>%
  filter(strata == "Race/ethnicity")

asthma_2020_white <- asthma_2020_race %>%
  filter(strata_name == "White")
asthma_2020_white <- asthma_2020_white[-(1),]
asthma_2020_white <- asthma_2020_white[order(asthma_2020_white$county),]
colnames(asthma_2020_white)[colnames(asthma_2020_white)== "county"] = "County" 
asthma_2020_white <- asthma_2020_white %>% mutate(County=paste(County,c("COUNTY"), sep=" "))%>%
  select(c("County","number_of_ed_visits")) %>%
  rename(white_asthma_ed_visits = "number_of_ed_visits")


asthma_2020_non_white <- asthma_2020_race %>%
  filter(strata_name != "White")
asthma_2020_non_white <- asthma_2020_non_white[-(1),]
asthma_2020_non_white <- asthma_2020_non_white[order(asthma_2020_non_white$county),]
asthma_2020_non_white <- asthma_2020_non_white %>%
  group_by(county) %>%
  summarise(num = n(),
            non_white = sum(number_of_ed_visits))
asthma_2020_non_white <- asthma_2020_non_white[-(6),]
colnames(asthma_2020_non_white)[colnames(asthma_2020_non_white)== "county"] = "County" 
asthma_2020_non_white <- asthma_2020_non_white %>% mutate(County=paste(County,c("COUNTY"), sep=" "))%>%
  select(c("County","non_white")) %>%
  rename(non_white_asthma_ed_visits = "non_white")


```

```{r Kathy 5}
#joining the scores with measures and stratified asthma datasets

CES_environment_Asthma <- scores_raw_county %>%
   left_join(measures_raw_county, by='County') %>%
   left_join(asthma_2020_age_adult, by="County") %>% 
   left_join(asthma_2020_age_child, by="County") %>%
   left_join(asthma_2020_white, by="County") %>%
   left_join(asthma_2020_non_white, by="County")


```

Data Dictionary
To foster collaboration and allow other colleagues to review our work, we created a new data dictionary with terms and definitions of variables used in our analysis.

```{r data dictionary - Kathy}

new_data_dictionary <- data.frame(
variables <- c('2020 Asthma ED visits, 2019 Census, and environmental burden summary','County', 'CES 4.0 Score', 'CES_score_quartile_range', 'Total Population', 'white_prct', 'white_pop',
               'non_white_prcp', 'non_white_pop', 'pop_under_10_prct', 'pop_under_10', 'pop_over_10_prct','pop_over_10',
               'Pollution Burden', 'Pollution Burden Score','Pollution Burden Pctl', 'Housing Burden', 'Housing Burden Pctl',
               'adult_asthma_ed_visits','child-asthma_ed_visits','white_asthma_ed_visits','non_white_asthma_ed_visits'), 
descriptions <- c('Data dictionary','California counties', 'CalEnviroScreen Score','CES 4.0 score grouped by quartile range of scores', '2019 ACS population estimates in census tracts', 'percetange of population that identify as white', 'estimated raw number of population that identify as white, calculated by multiplying percentage to total population', 'percentage of population that identify as non-white or multiple race, calculated by summming percentages of all non-white races and others/multile race from raw demographic results in census tract', 'estimated raw number of population that identify as non-white or multiple race, calculated by multiplying percentage to total population', 'percentage of population that is less than 10 years old','estimated raw number of population that is less than 10 years old, calculated by multiplying percentage to total populaiton', 'percentage of population that is more than 10 years old, calculated by summing population 10 to 64 years and over 64 years', 'estimated raw number of population that is more than 10 years old, calculated by multiplying percentage to total population','Average of percentiles from the Pollution Burden indicators (with a half weighting for the Environmental Effects indicators)','Pollution Burden variable scaled with a range of 0-10. (Used to calculate CES 4.0 Score)','Pollution burden percentile','Percent housing burdened low income households', 'Housing burden percentile','number of ED visits due to asthma for those over 18 years of age', 'number of ED visits due to asthma for those under 18 years of age', 'number of ED visits due to asthma for those that identify as white', 'number of ED visits due to asthma for those that identify as non-white or multiple race'))

colnames(new_data_dictionary)[1] = "VARIABLE"
colnames(new_data_dictionary)[2] = "DESCRIPTION"

  
```

Data Analysis
In this section, we looked at county-level CES scores, Pollution Burden Score, and Housing Burden Percentage. The CES score estimates the overall environmental and socioeconomic/population-related burden on a community, with the maximum score being 100; higher values mean a higher burden on the community. Pollution Burden Score is a numeric score that estimates the effects of environmental exposures and environmental effects on a community, with the score's value ranging from 0.1 to 10, inclusive. Housing Burden Percentage refers to the percentage of households that are both low income (making less than 80% of their county's median family income) and severely burdened by rent (paying greater than 50% of their income toward housing costs). Weighted averages were calculated and placed into a table named weighted_environment.


```{r CES Scores/Measures Analysis}

#3A: County and: CES score, Pollution Burden, Housing Burden %

#subsetting the combined dataset to analyze statistics for specific environmental measures by county 

asthma_house_poll <- data.frame("County" = CES_environment_Asthma$County, 
                                "CES 4.0 Score" = 
                                  CES_environment_Asthma$`CES 4.0 Score`, 
                                "Pollution Burden Score" = 
                                CES_environment_Asthma$`Pollution Burden Score`, 
                                "Housing Burden %" = 
                                  CES_environment_Asthma$`Housing Burden`, 
                                "Population" = 
                                  CES_environment_Asthma$`Total Population`)

colnames(asthma_house_poll) = c("County", "CES 4.0 Score", 
                                "Pollution Burden Score", 
                                "Housing Burden %", "Population")


#calculating the weighted averages for CES/pollution/housing

asthma_house_poll$weights <- 
  asthma_house_poll$`Population`/sum(asthma_house_poll$`Population`)

weighted_CES <-
  sum(asthma_house_poll$weights*asthma_house_poll$`CES 4.0 Score`)

weighted_pollution_burden <-
  sum(asthma_house_poll$weights*asthma_house_poll$`Pollution Burden Score`)

weighted_housing_burden <-
  sum(asthma_house_poll$weights*asthma_house_poll$`Housing Burden %`)

weighted_environment <- data.frame(weighted_CES, weighted_pollution_burden, weighted_housing_burden)

colnames(weighted_environment) = c("Weighted CES 4.0 Score", 
                                   "Weighted Pollution Burden", 
                                   "Weighted Housing Burden %")

```
We then looked at age and asthma rates, comparing the rates for adults and children, and then using the proportions to calculate the weighted state ED admission rates for adults and children. These calculations were then placed in a dataset, asthma_2020_age_final. 

```{r Age and Asthma Rates}

#joining the child and adult asthma ED admission datasets

asthma_2020_age_report <- asthma_2020_age_adult %>%
   left_join(asthma_2020_age_child, by='County')

#calculating the weighted state ED admission rate for adult and child
adult_total_ed_visits <- sum(asthma_2020_age_report$adult_asthma_ed_visits)
child_total_ed_visits <- sum(asthma_2020_age_report$child_asthma_ed_visits)

#defining a column that represents the proportion of adult/child ED admissions contributed by county 

asthma_2020_age_report$adult_visit_state_prop <- 
  asthma_2020_age_report$adult_asthma_ed_visits/adult_total_ed_visits 

asthma_2020_age_report$child_visit_state_prop <- 
  asthma_2020_age_report$child_asthma_ed_visits/child_total_ed_visits 

#using above proportion to calculate weighted state ED admission rates for adult and child
adult_weighted_state_ed_rate <- 
  sum(asthma_2020_age_report$adult_visit_state_prop*
        asthma_2020_age_report$`age-adjusted_ed_visit_rate.x`)

child_weighted_state_ed_rate <- 
  sum(asthma_2020_age_report$child_visit_state_prop*
        asthma_2020_age_report$`age-adjusted_ed_visit_rate.y`)

#cleaning up the calculation worksheet, asthma_2020_age_report
asthma_2020_age_final <- select(asthma_2020_age_report, 
                                "age-adjusted_ed_visit_rate.x", 
                                "age-adjusted_ed_visit_rate.y")
colnames(asthma_2020_age_final) = c("County",
                                    "Adult Age-Adjusted ED Visit Rate", 
                                    "Child Age-Adjusted ED Visit Rate")


```
Please see below for the final tables of our analyses.


```{r Statistics and Tables}
#joining the cleaned up CES data with the cleaned up age-asthma data in preparation for summary statistics and tables

data_final <- asthma_2020_age_final %>%
   left_join(asthma_house_poll, by='County')
data_final <- data_final[,-c(7,8)]

#summary statistics for the CES/age-asthma data
print(summary(data_final))

adult_quartile <- quantile(data_final$`Adult Age-Adjusted ED Visit Rate`, probs = c(0.25,0.5,0.75,1))
child_quartile <- quantile(data_final$`Child Age-Adjusted ED Visit Rate`, probs = c(0.25,0.5,0.75,1))
ces_quartile <- quantile(data_final$`CES 4.0 Score`, probs = c(0.25,0.5,0.75,1))
pollution_quartile <- quantile(data_final$`Pollution Burden Score`, probs = c(0.25,0.5,0.75,1))
housing_quartile <- quantile(data_final$`Housing Burden %`, probs = c(0.25,0.5,0.75,1))

quartiles <- data.frame(adult_quartile,child_quartile,ces_quartile,pollution_quartile,housing_quartile)
colnames(quartiles) = c("Quartiles of CA County Adult ED Admission Rate", "Quartiles of CA County Child ED Admission Rate", "Quartiles of CA County CES 4.0 Scores", "Quartiles of CA County Pollution Burden Scores", "Quartiles of CA County Housing Burden %")

#weighted average table
weighted_final <- data.frame(adult_weighted_state_ed_rate, 
                             child_weighted_state_ed_rate, 
                             weighted_CES, weighted_pollution_burden, 
                             weighted_housing_burden)

colnames(weighted_final) <-
  c("Adult Weighted State ED Admission Rate",
    "Child Weighted State ED Admission Rate",
    "Weighted State CES",
    "Weighted State Pollution Burden",
    "Weighted State Housing Burden %")

rownames(weighted_final) <- c("Weighted Value")

print(t(quartiles))
print(t(weighted_final))


```
We attempted to analyze the asthma ED visit rates by race but ran into issues due to the year mismatch between the datasets. We found that there were certain values where the non-white asthma visit rates were larger than the non-white population of the county. While this may be due to the fact that there can be more ED visit rates than people, we wanted to look more into this phenomenon to determine its accuracy and decide how it can be handled.


```{r Race}

#need some work on race for final submission

# #3C: obtain the race populations from CES_Environment_asthma, and divide the ED visits in asthma_2020 white and non-white by this population. note that we assume that the racial populations in each county don't change much year-over-year. This is a third table. column 1 is county, column 2 is racial group, column 3 is rate
# 
# #report out summary ed visits for: white, non-white
# 
# asthma_race_report <- select(CES_environment_Asthma, "County", "white_pop", "white_asthma_ed_visits", "non_white_pop", "non_white_asthma_ed_visits")
# 
# white_total_ed_visits <- sum(asthma_race_report$white_asthma_ed_visits)
# non_white_total_ed_visits <- sum(asthma_race_report$non_white_asthma_ed_visits)
# 
# 
# asthma_race_report$white_visit_state_prop <- asthma_race_report$white_asthma_ed_visits/white_total_ed_visits 
# 
# asthma_race_report$non_white_visit_state_prop <- asthma_race_report$non_white_asthma_ed_visits/non_white_total_ed_visits 
# 
# 
# asthma_race_report$white_asthma_ed_rate <- asthma_race_report$white_asthma_ed_visits/asthma_race_report$white_pop 
# asthma_race_report$non_white_asthma_ed_rate <- asthma_race_report$non_white_asthma_ed_visits/asthma_race_report$non_white_pop
# 
# white_weighted_state_ed_rate <- sum(asthma_race_report$white_visit_state_prop*asthma_race_report$white_asthma_ed_rate)
# 
# non_white_weighted_state_ed_rate <- sum(asthma_race_report$non_white_visit_state_prop*asthma_race_report$non_white_asthma_ed_rate)
# 
# asthma_race_final <- select(asthma_race_report, County, white_asthma_ed_rate, non_white_asthma_ed_rate)

```

Limitations
One limitation of our project is that the CES datasets included data from 2021, while the CHHS dataset included data from 2020. (CES does not have data for 2020, and CHHS does not have data for 2021.) This year mismatch may make our race-based asthma ED visit rate calculations a bit off due to the growth in population. We can possibly reduce the negative impacts by using formulas to transform CHHS' data from 2020 into 2021 population estimates. More work is needed to see if this is necessary or viable.
Another limitation is that we used averages for summing the census tract data into county-level data. This method is the easiest as it treats all census tracts as equal in terms of their contribution to county-level data, but does not provide a heavier weight for more populous census tracts. Therefore, census tracts with higher populations may be underestimated, and census tracts with lower populations may be overestimated.
A third limitation is that the "asthma" dataset's age groupings do not match 1:1 with the CES datasets. The "asthma" dataset has age strata such as 0-17, 18+, 0-4, 5-17, 18-64, 65+ while the CES datasets have <10, 10-64, and 65+. We can match the elderly strata across the datasets, but may need another method for looking at child/adolescent strata.

Conclusion
Our team welcomes any feedback on how we can improve our analysis of the Cal Enviro Screen data and CHHS' asthma visit rates. Thank you for your support.

``